Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-20 submission/manuscript.Rmd
First pass at reducing the number of References

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016]. Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Hannigan2017; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

SCFAs have have anti-inflammatory and anti-proliferative activities [@Encarnao2018; @Verma2018; @Zheng2017; @OKeefe2016]. Furthermore, manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018; @Tian2018]. These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Vital2014; @Sanna2019; @Liu2019; @Meisel2016]. Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Schatzkin2000; @Yao2017; @Kunzmann2015; @Murphy2012; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

@@

**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, isobutyrate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A). The median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas. The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`). The median concentration of isobutyrate in people with normal colons (`r iso_p_cross$normal` mmol/kg) was between those with adenomas (`r iso_p_cross$adenoma` mmol/kg) and carcinomas (`r iso_p_cross$cancer` mmol/kg). Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg (P=`r iso_p_follow$p.value`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we pooled the SCFA concentrations on a per molecule of carbon basis and again failed to see any significant differences (P>`r pooled_test`). The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to this observation.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs. Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B). Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using picrust2. Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.

@@

**Measuring specific SCFAs.** The measurement of acetate, propionate, isobutyrate, and butyrate used a previously published protocol that used High-Performance Liquid Chromatography (HPLC) [@Venkataraman2016]. Two changes were made to the protocol. First, instead of using fecal samples suspended in DNA Genotek OmniGut tubes, we suspended frozen fecal samples in 1 mL of PBS. Second, instead of using the average weight of fecal sample aliquots to normalize SCFA concentrations, we used the actual weight of the fecal samples. These methodological changes did not affect the range of concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).

**16S rRNA gene sequence data analysis.** Sequence data from Baxter et al. [@Baxter2016] and Sze et al. [@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009]. The sequence generation workflow and processing steps have been previously described [@Schloss2009; @Kozich2013]. In brief, the original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer [@Kozich2013]. The resulting sequence data were assembled into contigs and screened for quality aberrations including chimeras, which were removed based on the use of UCHIME [@Edgar2011]. Sequences were clustered into OTUs at 97% similarity using OptiClust [@Westcott2017] and assigned to the closest possible genus using a naive Bayesian classifier with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16) [@Wang2007]. We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].

**Metagenomic DNA sequence analysis.** A subset of the samples from the samples described by Baxter et al. [@Baxter2016] were used to generate metagenomic sequence data (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`). These data were generated by Hannigan et al. [@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915). Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads. The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data. We downloaded the sequences and assembled them into contigs using MEGAHIT [@Li2015], which were used to identify open reading frames (ORFs) using Prodigal [@Hyatt2010]. We determined the abundance of each ORF by mapping the raw reads back to the ORFs using Diamond [@Buchfink2014]. We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017; @Schloss2008]. We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped [@Kanehisa2015].

>>>>

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis. Numerous studies in murine models have supported this model [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

SCFAs have have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018]. These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

@@

**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, isobutyrate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A). Interestingly, the median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas, respectively. The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`). Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg (P=`r iso_p_follow$p.value`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we pooled the SCFA concentrations on a per molecule of carbon basis and again failed to see any significant differences (P>`r pooled_test`). The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to this observation.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs. Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B). Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using PICRUSt. Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.

@@

**Measuring specific SCFAs.** The measurement of acetate, propionate, isobutyrate, and butyrate used a previously published protocol that used High-Performance Liquid Chromatography (HPLC) [@Venkataraman2016]. Two changes were made to the protocol. First, instead of using fecal samples suspended in DNA Genotek OmniGut tubes, we suspended frozen fecal samples in 1 mL of PBS. Second, instead of using the average weight of fecal sample aliquots to normalize SCFA concentrations, we used the actual weight of the fecal samples. These methodological changes did not affect the range of concentrations of these SCFAs between the two studies.

**16S rRNA gene sequence data analysis.** Sequence data from Baxter et al. [@Baxter2016] and Sze et al. [@Sze2017] were obtained from the Sequence Read Archive (studies SRP062005 and SRP096978) and reprocessed using using mothur v.1.42 [@Schloss2009]. The original studies generated sequence data from V4 region of the 16S rRNA gene using paired 250 nt reads on an Illumina MiSeq sequencer. The resulting sequence data were assembled into contigs, screened to remove low quality contigs and chimeras. The curated sequences were then clustered into OTUs at a 97% similarity threshold and assigned to the closest possible genus with an 80% confidence threshold trained on the reference collection from the Ribosomal Database Project (v.16). We used PICRUSt (v.2.1.0-b) with the recommended standard operating protocol to generate imputed metagenomes based on the expected metabolic pathways, KEGG categories, and enzyme commission numbers [@Langille2013].

**Metagenomic DNA sequence analysis.** A subset of the samples from the samples described by Baxter et al. [@Baxter2016] were used to generate metagenomic sequence data (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`). These data were generated by Hannigan et al. [@Hannigan2017] and deposited into the Sequence Read Archive (study SRP108915). Fecal DNA was subjected to shotgun sequencing on an Illumina HiSeq using 125 bp paired end reads. The archived sequences were already quality filtered and aligned to the human genome to remove contaminating sequence data. We downloaded the sequences and assembled them into contigs using MEGAHIT [@Li2015], which were used to identify open reading frames (ORFs) using Prodigal [@Hyatt2010]. We determined the abundance of each ORF by mapping the raw reads back to the ORFs using Diamond [@Buchfink2014]. We clustered the ORFs into operational protein families (OPFs) in which the clustered ORFs were more than 40% identical to each other using mmseq2 [@Steinegger2017]. We also used mmseq2 to map the ORFs to the KEGG database and clustered the ORFs according to which category the ORFs mapped.
